---
title: "delta downscaling"
output: rmarkdown::pdf_document
#output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{delta downscaling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# THIS IS WORK IN PROGRESS, DO NOT USE FOR ANY ANALYSIS

# THIS IS A NEW VERSION OF THE VIGNETTE FOCUSSIN ON EUROPE SO THAT WE CAN SEE
# HOW ICE SHEETS ARE DOWNSCALED

Note that we will use the draft functions for downscaling, which can be accessed
by using the prefix `pastclim:::`. The `:::` accesses internal functions. You will
also get a message reminding you that these functions are not final and thus
should not be used for real analysis.


## Delta downscaling a dataset in `pastclim`

The `terra` package can downscale reconstructions using a bilinear method. Other `R`
packages also offer various approaches to downscale rasters.

For palaeoclimate reconstructions, the delta method has been shown to be very effective
(Beyer et al, REF), but sea level changes require some care on how to apply such a method.
`pastclim` includes functions to use the delta method for downscaling. We will focus on South East
Asia, as this area was greatly affected by sea level changes over the last glaciation. Specifically, we will look at the area around Borneo (for real applications, we would
reccomend using a bigger extent in areas of large changes in land extent, as
interpolating over a small extent can lead to greater artefacts; for this example,
we keep the extent small to reduce computational time).

## An example for one variable

When downscaling, it is generally a better idea to downscale the monthly variables,
and then recompute the BIOCLIM variables. We will use the monthly variables
from the Beyer2020 dataset, and focus on the time steps that are also available
in the Example dataset (i.e. a subset of the full Beyer2020 dataset, to reduce
computational time):
```{r}
library(terra)
library(pastclim)
europe_ext <- pastclim::region_extent$Europe
tavg_vars <- c(paste0("temperature_0",1:9),paste0("temperature_",10:12))
time_steps <- get_time_steps(dataset = "Example")
```
The series for this region can be generated with the following command (note
that you will need the appropriate variables for the Beyer2020 dataset downloaded
on your computer, use `download_dataset` if needed):
```{r eval=FALSE}
tavg_series <- region_series(bio_variables =tavg_vars,
                             time_bp =  time_steps,
                             dataset = "Beyer2020",
                              ext = europe_ext)
```

```{r echo=FALSE, results="hide"}
#tavg_series <- terra::sds(system.file("extdata/delta/tavg_series.nc",
#                                       package="pastclim"))

```

Downscaling is performed one variable at a time. We will start with temperature in January.
So, we first need to extract the `SpatRaster` of model low resolution data from the `SpatRasterDataset`:
```{r}
tavg_model_lres_rast <- tavg_series$temperature_01
tavg_model_lres_rast
```
And we can now plot it:
```{r, fig.width=6, fig.height=5}
plot(tavg_model_lres_rast, main = time_bp(tavg_model_lres_rast))
```

We can see how that the reconstructions are rather coarse (the Beyer2020 dataset
uses 0.5x0.5 degree cells). We now need a set of
high resolutions observations for the variable of interest that we will use to 
generate the delta raster used to downscale reconstructions. We will use data from
WorldClim2 at 10 minute resolution (but other datasets such as CHELSA would be
equally suitable). For WorldClim, there are some convenient functions to download
and then load variables:
```{r eval=FALSE}
pastclim:::download_worldclim("tavg",10)
```
Once the variable is downloaded, we can load it at any time with:
```{r eval=FALSE}
tavg_obs_hres_all<- pastclim:::load_worldclim("tavg",10)
```

```{r echo=FALSE, results="hide"}
#tavg_obs_hres_all<- terra::rast(system.file("extdata/delta/tavg_obs_hres_all.nc",
#                                            package="pastclim"))
```

We want to crop these reconstructions to the extent of interest
```{r, fig.width=4, fig.height=4}
tavg_obs_hres_all <- terra::crop(tavg_obs_hres_all, europe_ext)
# extract the January raster
tavg_obs_hres_rast <- tavg_obs_hres_all[[1]]
plot(tavg_obs_hres_rast)
```

We need to make sure that the extent of the modern observations is the same as the
extent of the model reconstructions:

```{r}
ext(tavg_obs_hres_rast)==ext(tavg_model_lres_rast)
```

If that was not the case, we would use `terra::crop` to match the extents.

We also need a high resolution global relief map (i.e. integrating both 
topographic and bathymetric values) to reconstruct past
coastlines following sea level change. The relief raster will need to have
the same extent and resolution as the high resolution observations. We
can download one (based on ETOPO2022) with:

```{r eval=FALSE}
relief_rast <- pastclim:::download_relief(tavg_obs_hres_rast)

```

```{r echo=FALSE, results="hide"}
#relief_rast <- terra::rast(system.file("extdata/delta/relief.nc",
#                                       package="pastclim"))
```

We can quickly confirm that the resulting relief raster has the same extent 
as the model reconstructions,
as well as the same resolution of the high resolution climate observations:

```{r}
ext(relief_rast) == ext(tavg_model_lres_rast)

ext(relief_rast) == ext(tavg_obs_hres_rast)
ncol(relief_rast) == ncol(tavg_obs_hres_rast)
nrow(relief_rast) == nrow(tavg_obs_hres_rast)
```

We can now generate a high resolution land mask for the periods of interest. By
default, we use the sea level reconstructions from Spratt et al 2016, but a different
reference can be used by setting sea levels for each time step (see the man page
for `make_land_mask` for details):

```{r, fig.width=6, fig.height=5}
high_res_mask <- pastclim:::make_land_mask(relief_rast = relief_rast, 
                                time_bp = time_bp(tavg_model_lres_rast))
plot(high_res_mask, main=time_bp(high_res_mask))
```

We can now compute a delta raster and use it to downscale the model
reconstructions:

```{r}
delta_rast<-pastclim:::delta_compute(x=tavg_model_lres_rast, ref_time = 0, 
                                     obs = tavg_obs_hres_rast)
model_downscaled <- pastclim:::delta_downscale (x = tavg_model_lres_rast, 
                                                delta_rast = delta_rast,
                                                x_landmask_high = high_res_mask)
model_downscaled
```

Let's inspect the resulting data:
```{r, fig.width=6, fig.height=5}
plot(model_downscaled, main = time_bp(model_downscaled))
```

Note the artefacts in Scandinavia at -15000 and -20000 years BP. This is due to the fact that, in the original
dataset, we have icesheets on those areas, so we don't have any information and end up
extrapolating over a huge area. The correct step then is to readd the icesheets. We start by getting
the ice sheets from the original dataset

```{r}
ice_mask_low_res <- get_ice_mask(time_bp=time_steps,dataset="Beyer2020")
ice_mask_low_res_bin <- pastclim:::make_binary_mask(ice_mask_low_res)
ice_mask_high_res <- resample(ice_mask_low_res_bin, high_res_mask, method="bilinear")
gf <- focalMat(ice_mask_high_res, c(0.5,3), "Gauss")
ice_mask_gf <- focal(ice_mask_high_res, w = gf)
ice_mask_gf[ice_mask_gf>0.5]<-NA
ice_mask_gf[!is.na(ice_mask_gf)]<-1
model_downscaled_masked <- mask(model_downscaled,ice_mask_gf)
plot(model_downscaled_masked)
```


And, as a reminder, the original reconstructions (note that the colour scales are not
the same! `terra` chooses a scale for each time step based on the time specific range):
```{r, fig.width=6, fig.height=5}
plot(tavg_model_lres_rast, main = time_bp(tavg_model_lres_rast))
```

## Computing the bioclim variables
To compute the bioclim variables, we need to repeat the procedure above for
temperature and precipitation for all months. Let us start with temperature. We
loop over each month, create a `SpatRaster` of downscaled temperature, add it to
a list, and finally convert the list into a `SpatRasterDataset`

```{r}
tavg_downscaled_list<-list()
for (i in 1:12){
  delta_rast<-pastclim:::delta_compute(x=tavg_series[[i]], ref_time = 0, 
                                       obs = tavg_obs_hres_all[[i]])
  tavg_downscaled_list[[i]] <- pastclim:::delta_downscale (x = tavg_series[[i]], 
                                                  delta_rast = delta_rast,
                                                  x_landmask_high = high_res_mask)
}
tavg_downscaled <- terra::sds(tavg_downscaled_list)
```

Quickly inspect the resulting dataset:
```{r}
tavg_downscaled
```
As expected, we have 12 months (subdatasets), each with 5 time steps.

We now need to create a series for precipitation. As precipitation is bound at 0, it is often downscaled by
using a log transformation (using a "multiplicative" rather than "additive" delta).
So, we use a *log(x+1)* transformation (the *+1* deals with the zeroes):
```{r eval=FALSE}
prec_vars <- c(paste0("precipitation_0",1:9),paste0("precipitation_",10:12))
prec_series <- region_series(bio_variables = prec_vars,
                             time_bp =  time_steps,
                             dataset = "Beyer2020",
                             ext = sea_ext)
prec_series <- sds(lapply(prec_series,FUN =function(i)(log(i+1))))
```

```{r echo=FALSE, results="hide"}
prec_vars <- c(paste0("precipitation_0",1:9),paste0("precipitation_",10:12))
prec_series <- terra::sds(system.file("extdata/delta/prec_series.nc",
                           package="pastclim"))
prec_series <- sds(lapply(prec_series,FUN =function(i)(log(i+1))))
```

Get some high resolution observations:
```{r, eval=FALSE}
pastclim:::download_worldclim("prec",10)
prec_obs_hres_all <- pastclim:::load_worldclim("prec",10)
prec_obs_hres_all <- terra::crop(prec_obs_hres_all, sea_ext)
prec_obs_hres_all <- log(prec_obs_hres_all+1)
```

```{r echo=FALSE, results="hide"}
prec_obs_hres_all <-terra::rast(system.file("extdata/delta/prec_obs_hres_all.nc",
                                            package = "pastclim"))
prec_obs_hres_all <- log(prec_obs_hres_all+1)
```

And finally downscale precipitation (note that we backtransform it with `exp(x)-1`. 
```{r}
prec_downscaled_list<-list()
for (i in 1:12){
  delta_rast<-pastclim:::delta_compute(x=prec_series[[i]], ref_time = 0, 
                                       obs = prec_obs_hres_all[[i]])
  prec_downscaled_list[[i]] <- exp(pastclim:::delta_downscale (x = prec_series[[i]], 
                                                  delta_rast = delta_rast,
                                                  x_landmask_high = high_res_mask))-1
}
prec_downscaled <- terra::sds(prec_downscaled_list)
```

We are now ready to compute the bioclim variables:
```{r}
bioclim_downscaled<-bioclim_vars(tavg = tavg_downscaled, prec = prec_downscaled)
```

Let's inspect the object:
```{r}
bioclim_downscaled
```

And plot the first variable (bio01):
```{r}
plot(bioclim_downscaled[[1]], main = time_bp(bioclim_downscaled[[1]]))
```

We can now save the downscaled `sds` to a netcdf file:
```{r}
terra::writeCDF(bioclim_downscaled,paste0(tempdir(),"/EA_bioclim_downscaled.nc"), overwrite=TRUE)
```

And then use it as a custom dataset for any function in `pastclim`. Let's extract
a region series for three variables:
```{r}
custom_data <- region_series(bio_variables =c("bio01","bio04","bio19"),
                             dataset = "custom",
                             path_to_nc = paste0(tempdir(),"/EA_bioclim_downscaled.nc"))
```

We can quickly inspect the resulting `sds` object:
```{r}
custom_data
```

And plot it:
```{r}
plot(custom_data$bio01,main=time_bp(custom_data$bio01))
```