---
title: "delta downscaling"
output: rmarkdown::pdf_document
#output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{delta downscaling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# THIS IS WORK IN PROGRESS, DO NOT USE FOR ANY ANALYSIS

Note that we will use the draft functions for downscaling, which can be accessed
by usign the prefix `pastclim:::`. The `:::` access internal functions. You will
also get a message reminding you that these functions are not final and thus
should not be used for real analysis.


## Downscaling a dataset in pastclim `pastclim`

The `terra` package can downscale reconstructions using a bilinear method. Other `R`
packages also offer various approaches to downscale rasters.

For palaeoclimate reconstructions, the delta method has been shown to be very effective
(Beyer et al, REF), but sea level changes require some care on how to apply such a method.
`pastclim` includes functions to use the delta method for downscaling. We will focus on South East
Asia, as this area was greatly affected by sea level changes over the last glaciation.
We will work on the variable "bio01". However, note that, in most cases, it is better practice to
downscale monthly temperature and precipitation estimates, and then compute the "bio" variables
from those quantities.

First, we extract a time series for the variable and region of interest:
```{r}
library(terra)
library(pastclim)
sea_ext<- terra::ext(100, 130, -15, 20)
model_series <- region_series(bio_variables = "bio01", dataset = "Example",
                            ext = sea_ext)
model_series
```

Note that a `SpatRasterDataset` has been returned, which contains only one `SpatRaster`
for "bio01". We want to work on that `SpatRaster`, so we extract it with:

```{r}
model_rast <- model_series$bio01
model_rast
```
And we can now plot it:
```{r, fig.width=6, fig.height=5}
plot(model_rast, main = time_bp(model_rast))
```

We can see how that the reconstructions are rather coarse (the Example dataset
uses 1x1 degree cells). We now need a set of
high resolutions observations for the variable of interest that we will use to 
generate the delta raster used to downscale reconstructions. We will use data from
Worldclim2 at 5 minute resolution (but other datasets such as CHELSA would be
equally suitable).
```{r, fig.width=4, fig.height=4}
high_res_file <- system.file("extdata/wc2.1_5m_bio_1_sea.tif",
                             package = "pastclim")
high_res_obs <- terra::rast(high_res_file)
plot(high_res_obs)
```

We need to make sure that the extent of the modern observations is the same as the
extent of the model reconstructions:

```{r}
ext(high_res_obs)==ext(model_rast)
```

If that was not the case, we would use `terra::crop` to match the extents.

We also need a high resolution global relief map (i.e. integrating both 
topographic and bathymetric values) to reconstruct past
coastlines following sea level change. The relief raster will need to have
the same extent and resolution as the high resolution observations. We
can download one (based on ETOPO2022) with:

```{r}
relief_rast <- pastclim:::download_relief(high_res_obs)

```

```{r eval=FALSE, results="hide",echo=FALSE}
relief_file <- system.file("extdata/ETOPO1_Ice_c_gmt4_5min_sea.tif",
                           package = "pastclim")
relief_rast <- terra::rast(relief_file)
```

We can quickly confirm that the resulting relief raster has the same extent 
as the model reconstructions,
as well as the same resolution of the high resolution climate observations:

```{r}
ext(relief_rast) == ext(model_rast)

ext(relief_rast) == ext(high_res_obs)
ncol(relief_rast) == ncol(high_res_obs)
nrow(relief_rast) == nrow(high_res_obs)
```

We can now generate a high resolution land mask for the periods of interest. By
default, we use the sea level reconstructions from Spratt et al 2016, but a different
reference can be used by setting sea levels for each time step (see the man page
for `make_land_mask` for details):

```{r, fig.width=6, fig.height=5}
high_res_mask <- pastclim:::make_land_mask(relief_rast = relief_rast, 
                                time_bp = time_bp(model_rast))
plot(high_res_mask, main=time_bp(high_res_mask))
```

We can now compute a delta raster and use it to downscale the model
reconstructions:

```{r}
delta_rast<-pastclim:::delta_compute(x=model_rast, ref_time = 0, 
                                     obs = high_res_obs)
model_downscaled <- pastclim:::delta_downscale (x = model_rast, 
                                                delta_rast = delta_rast,
                                                x_landmask_high = high_res_mask)
model_downscaled
```

Let's inspect the resulting data:
```{r, fig.width=6, fig.height=5}
plot(model_downscaled, main = time_bp(model_downscaled))
```

And, as a reminder, the original reconstructions (note that the colour scales are not
the same! `terra` chooses a scale for each time step based on the time specific range):
```{r, fig.width=6, fig.height=5}
plot(model_rast, main = time_bp(model_rast))
```
